Sea-ice and macrozooplankton distribution as determinants of top predator community structure in Antarctic winter

### 2024-05-30 analysis/paper/paper.qmd
paper: methods

# Background

# Methods

>>>>

# Materials and Methods

## Survey area data

### Survey design

We surveyed the South Shetland Island and Elephant Island regions during the austral winters (August to September) of 2012–2016 as part of the US Antarctic Marine Living Resources (AMLR) program aboard the RV Ice Breaker (RVIB) ‘Nathaniel B. Palmer’ (Fig. 1). The main objectives of the survey were to map hydrographic conditions, study the distribution and abundance of fish, krill, and other zooplankton through acoustic methods and net hauls, and describe the spatial distribution of seabirds and marine mammals. We followed methods consistent with the long-term austral summer surveys conducted by the US AMLR program [@reissOverwinterHabitatSelection2017]. We sampled between 33 and 108 stations per year, dependent on ice conditions. Stations were spaced approximately 25–30 km apart, and the full survey area covered an area of approximately 120000 km^2^.

### Visual surveys

Sighting surveys for seabirds and marine mammals were conducted from the bridge of the ship, following established methodologies described in previous studies [@santoraEnvironmentalDeterminantsTop2014]. Two observers collected observations simultaneously during daylight hours. One observer recorded seabirds, while the other scanned for marine mammals in the water and on the ice using binoculars. Observers used 10 × 50 binoculars and image-stabilized 20 × 60 binoculars. The pilot house/bridge of the RVIB ‘Nathaniel B. Palmer’ is 54 ft (\~16.5 m) above sea level. Observations only stopped when the ship could not move forward though the sea ice, and no observations were made during station sampling.

During each survey, we continuously monitored and classified sea-ice conditions according to the international sea ice protocol (ASPeCt) [@worbyPartObservationalTechnique1999; @santoraSpatiotemporalPersistenceTop2013; @reissOverwinterHabitatSelection2017]. This involved estimating the percentage of ice cover on a scale of 0 to 10, where 0 is open water and 10 is complete ice coverage (fully consolidated). We also identified different types of ice, including first-year, multi-year, and thin gray ice. An automatic logger recorded the ice conditions every 10 s, updated by the observer as conditions changed.

### Hydrology

We measured hydrology at each survey station. Conductivity and temperature in the water column were measured using a Sea-Bird Inc. SBE-9/11+ CTD profiler (Table 1). We used Niskin sampling bottles to collect water samples for measuring chlorophyll *a* (chl *a*) and phaeopigment concentrations (Table 1). For full details, see @dietrich2021winter.

## Statistical analysis

We standardized our data to integrate visual surveys (line transects) with oceanographic sampling stations (discrete points). First, we aggregated all transect observations (relative abundance counts of seabirds and mammals, and ice conditions) into 1.85 km bins. We then associated each sampling station with all transect intervals conducted within 15 km and 3 d of the station to form ‘sites’. We excluded intervals farther in space or time than 15 km and 3 d from a sample station from the community analysis. After associating transect intervals and sample stations in sites, we aggregated the relative predator abundance and ice conditions of each site. We normalized relative predator abundance by sampling effort (ind. km^–1^), then applied a pseudo-log transformation (i.e. log~10~(*x* + 1)) to minimize the influence of highly abundant taxa. The analysis of macrozooplankton in these surveys [@dietrich2021winter] excluded taxa observed at fewer than 5% of sample stations; the seabird and marine mammal assemblage is much less diverse than macrozooplankton, so we did not exclude any species. We aggregated ice conditions across intervals within a site as the most frequent ice type (open, thin, first-year, or multi-year sea ice) and mean percent coverage.

We assessed the community structure from relative abundance data using cluster and ordination analyses. First, we calculated a Bray-Curtis dissimilarity matrix using the R package 'vegan' [@oksanenVeganCommunityEcology2022] to estimate dissimilarities in species composition between sites. We then generated hierarchical clusters using Ward's minimum variance linkage method, implemented in the R package 'cluster' [@maechlerClusterClusterAnalysis2022], applied to the dissimilarity matrix. We determined the optimal number of clusters (*k*) to retain from the hierarchy using the gap statistic [@tibshiraniEstimatingNumberClusters2001]. The gap statistic compares the within-cluster variation of different values of *k* to their expected values under a null reference distribution. The gap statistic, *f*, usually increases with *k*, but exhibits a shoulder at the optimal number of clusters. A recommended heuristic for finding the shoulder is selecting the smallest *k* such that the gap statistic for *k*, *f*(*k*), is greater than the gap statistic for *k + 1* less the standard error, *s* [@eq-gap-stat].

$$
f(k)>f(k+1)-s(k+1)
$$ {#eq-gap-stat}

We described each predator cluster with a smaller number of indicator species. Using the R package 'labdsv', we calculated indicator values, *d*, which indicate the relative affinity of a species to a particular cluster [@dufrene1997species; @robertsLabdsvOrdinationMultivariate2019]. The indicator value for species *i* in cluster *c*, *d~ic~*, is the product of relative frequency, *f~ic~*, and relative abundance, *a~ic~*, where *f~ic~* is the fraction of sites in *c* where *i* was present and *a~ic~* is the ratio of per-site abundance of *i* in *c* to the overall per-site abundance of *i*. *d* ranges between 0 and 1, where values close to 1 indicate a species that was ubiquitous within the cluster and highly abundant relative to other clusters. Indicator values ≥0.25 were retained as significant indicators.

To explore the relationship between community structure and the environment, we reduced the dimensionality of the community data using non-metric multidimensional scaling (NMDS). For this purpose, we used the same Bray-Curtis dissimilarity matrix from the cluster analysis. We determined the number of NMDS axes to retain by examining a scree plot, which plots overall stress of the fit against the number of axes, and assessed quality of fit using a Shepard diagram, which plots the dissimilarity distances in the reduced space against the original dissimilarity distances. We chose the smallest number of axes with stress \<0.2 and examined the correlation in the Shepard diagram for goodness of fit.

We used linear regression to quantify the relationship between the retained NMDS axes and the environmental, biological, and temporal variables. We performed this analysis with the `envfit()` function in the 'vegan' package, which treats environmental variables as the response variables and NMDS ordination scores as the predictors. The environmental variables included physical conditions (hydrology and ice conditions), biology (macrozooplankton community), and temporal variables (Table 1). Temporal variables included time of day of the net tow, to account for diel patterns in macrozooplankton distribution, as well as year, to account for interannual variability. The regression coefficients obtained from the analysis represent the direction and magnitude of the maximal change in community structure associated with each environmental variable. We tested for significant associations between predator and macrozooplankton communities using a chi-squared test and assessed the statistical significance of individual predator–prey community associations via post hoc analysis of the Pearson residuals.
